The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at ESO/VLT under programme 088.A-0051 and 091.D-0904, at TNG under programme A24TAC\_38, at Gemini North under programme GN-2011B-Q-10 and GTC under programme GTC43-11B.}

### 2017-11-25 referee_reply.txt

--1st sentence: Define sGRB, NS. 
JS: Done.

--For detection of kilonovae, also cite Berger+ 2013. 
JS: Definitely.

--"duration T_90 < 2" - make sure to qualify with 2 sec
JS: Done.

--Here and throughout, there is a tendency to use sGRB/short GRB and lGRB/long
GRB. I suggest using one or the other throughout the text. 
JS: We have streamlined this a bit, but have kept the wording "long and short GRBs"

@@

--"This number is 3.2 Gyr for the next-highest spectroscopic redshift" - This
sentence needs a citation and would be helpful to state what burst this is
JS: We have expanded this entire section a bit to be more complete.

--Even though the offset does not change much, it is worth recalculating in this
paper for completeness.
JS: Done

--"Along with the absence of dust" -- This is not yet proven in this paper. I
suggest revising to "If the assumption of no dust is correct..."
JS: This has been changed to "low dust content"

--The scenario in which the density of material is high to explain the high NH
would imply a bright afterglow, although would appear faint at z=2.211. Are the
afterglow limits still consistent with a high-density environment at z=2.211?
JS: The X-ray afterglow is not strongly dependent of the density and because of the optical
darkness, dust extinction locally to the GRB can accomodate a large brightness
interval. We have mentioned this in text.

@@

Fig. 4: A comparison of the light curve for GRB 111117A to the identical figure
in Margutti+ (their Fig 6) looks like the luminosity does not change, even
though it should be significantly higher given the revised redshift. In fact,
the Margutti+ light curve exceeds 10^48 erg s^-1 near the start, while the light
curve in Fig. 4 does not. Is the luminosity calculated correctly, or are they in
different energy bands? -Does this include the Chandra data as well? If so, it
would be good to include that in the caption.

Margutti+ didn't use the light curve from the Swift Burst Analyser. Her early light
curve is ~0.18 dex brighter. The shape is also a bit different. k-correction:
Margutti+ fixed the photon index to 2.2. The Burst Analyser provides estimates of
the photon index for each observation. At the time of the first observation, the
Burst Analyser gives a photon index of 0.93. The k-correction for a power-law
shaped spectrum is: (1+z)^(Gamma-2). The difference in these parameters
corresponds to a difference in the k-correction of 0.63 dex. The different
distances gives a luminosity difference of 0.58 dex (differences in the
cosmology are taken into account).

### 2017-11-25 GRB111117A.tex
Last edits before resubmission

We use the spectroscopic redshift, as well as new imaging data to constrain the
nature of the host galaxy and the physical parameters of the GRB. The rest-frame
X-ray derived hydrogen column density, for example, is the highest compared to a
complete sample of sGRBs and seems to follow the evolution with redshift as
traced by the hosts of long GRBs (lGRBs).

The host lies in the brighter end of the expected sGRB host brightness
distribution at $z = 2.211$, and is actively forming stars. Using the host as a
benchmark for redshift determination, we find that between 43 and 71 per cent of
all \textit{Swift}-detected sGRBs are too faint at $z\sim2$ to allow for a secure
redshift determination. The high redshift of GRB~111117A is evidence against a
lognormal delay-time model for sGRBs through the predicted redshift distribution
of sGRBs, which is very sensitive to high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts (sGRBs) come
from the merger of neutron stars (NSs), either with another NS, or a black hole,
due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Berger2013b, Yang2015, Jin2016, Rosswog2016}. This association was
recently confirmed by the simultaneous and co-spatial detection of gravitational
waves (GWs) from binary neutron star merger and a sGRB
\citep{LIGOScientificCollaboration2017a, Goldstein2017, Savchenko2017}. The
absence of associated supernovae in deep searches
\citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b, Kann2011} supports this idea and
distinguishes the physical origin of sGRBs from their long-duration
counterparts, \citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}.

The classification of GRBs in two groups initially comes from the bimodal
distribution of burst duration and spectral hardness in the BATSE sample \citep{Kouveliotou1993},
where the duration T$_{90} < 2$ sec has been regarded as the dividing line between long
and short GRBs. Additionally, it has been found for long GRBs (lGRBs) that there is a
spectral lag in the arrival-time of photons, with the most energetic ones
arriving first. This lag is consistent with zero for sGRBs
\citep{Norris2006}. Because both populations have continuous, overlapping
distributions in their observables and because telescopes observe in differing
bands, it is difficult to impose a single demarcation criterion between the two
classes. For this reason, the distinction between long and short GRBs is
preferably based on a combination of the high-energy properties \citep{Zhang2009,
	Kann2011, Bromberg2012a, Bromberg2013}.

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}, but because
most of these measurements come from the associated hosts, it is potentially
biased towards lower redshifts. Additionally, because the \textit{Swift}
sensitivity to sGRB decreases sharply with redshift \citep{Behroozi2014}, the
intrinsic redshift distribution of sGRBs is largely unknown at higher redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above the detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long lived in comparison to lGRBs, with the host stellar mass affecting
the sGRB rates more than star-formation rate (SFR) \citep{Berger2014}.

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers, which holds the promise for a detection of an associated
gravitational wave (GW) signal \citep{Ghirlanda2016}. This epochal breakthrough
occurred recently when the first NS-NS GW event ever was detected by LIGO/Virgo
(GW 170817) and associated to the weak sGRB 170817A detected by the Fermi and
Integral satellites \citep{LIGOScientificCollaboration2017a, Goldstein2017,
	Savchenko2017}. The simultaneous detection of a sGRB and a GW provides new
promising ways to constrain the binary inclination angle \citep{Arun2014, LIGOScientificCollaboration2017a} and
measure cosmological distances \citep{Nissanke2010, LIGOScientificCollaboration2017c}.

The total lifetimes of NS binaries depends on their orbit, mass, spin, initial
separations and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated using host galaxy star-formation history models and
spatial distribution of sGRBs within their hosts \citep[][]{Berger2014}. The
most distant cosmological bursts, however, offer direct, hard limits on the
coalescence time scales.

We here present a spectrum of the host galaxy of the short
GRB~111117A ($T_{90}=0.46$~s) and measure its redshift to be $z=2.211$. This
value is significantly higher than the previously estimated redshift based on
photometric studies \citep{Margutti2012,Sakamoto2013}. We present the GRB's rest
frame properties based on this new distance compared to previous analyses and
revisit the host properties derived from the new solution to the spectral energy distribution (SED) fit.
While no optical afterglow was detected, the excellent localization from a
detection of the X-ray afterglow by the \emph{Chandra X-ray Observatory}
allows us to discuss the positioning and environmental properties of this
remarkably distant sGRB.
We use the $\Lambda$CDM cosmology parameters provided by
\citet{Planck2015} in which the universe is flat with $H_0 = 67.7$\,km\,s$^{-1}$
and $\Omega_m = 0.307$. All magnitudes are given in the AB system.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{
	FORS2 R-band imaging of the field of GRB~111117A with the X-shooter slit overlaid. The slit
	position represents 4 epochs of spectroscopic observations taken at similar
	position angles. The corresponding photometry
	is shown in Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position as
	derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. 
	}
	\label{fig:spec_setup}
\end{figure}

@@

\begin{figure*}
	\centering
	\includegraphics[width=18cm]{figures/grid.pdf}
	\caption{2D-images of the emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959, \oiii$\lambda$5007, and \ha. The location of a bright skylines are marked by the blue boxes. The location of the emission line is indicated with the red ellipse. Because the host is observed in nodding-mode, negative images of the emission line appear on both sides in the spatial direction. For the upper, left panel containing \lya, the systemic redshoft position of \lya is marked by a yellow, vertical, dashed line. The red part of the \oii$\lambda$3727-doublet is affected by atmospheric absorption.}
	\label{fig:line}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/SEDspecphot.pdf}
	\caption{SED fit showing the best-fit SED to the derived photometry. The detection of \lya{}  is predicted from the SED fit and confirmed by the spectroscopic observations. Overplotted in grey is the observed spectrum, binned by 6 \AA{} for presentation purposes. An attempt to correct for slit losses has been carried out based on the average seeing of the observations. The blue, dashed line is the corresponding error spectrum, smoothed for presentation purposes. The reason for the spectral gaps at 5500 \AA{} and 10000 \AA{} is from the merging of the arms.}
	\label{fig:SED}
\end{figure*}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). Observations use a simple ABBA nodding pattern, with 5
arcsec nod throws. X-shooter covers the wavelength range from 3000~\AA{} to
24\,800~\AA{} (21\,000~\AA{} when the $K$-band blocking filter is used) across
three spectroscopic arms. We carried out the bias-correction, flat-fielding,
order tracing, wavelength calibration, rectification, and flux calibration using
the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} run
in physical mode. Because the echelle orders are curved across each detector, a
rectification algorithm is employed which introduces correlations between
neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the
UVB/VIS/NIR arm to minimize the degree of correlation while conserving the
maximal resolution. The observations are combined and extracted using scripts
described in Selsing et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}},
where the full spectral point spread function is modeled across each arm and
used for the optimal extraction algorithm \citep{Horne1986}. An overview of the
spectroscopic observations is given in Table~\ref{tab:spec_overview}, and the
slit position is shown in Fig.~\ref{fig:spec_setup}. Each of the epochs are
extracted individually and combined in a weighted fashion where the weights at
each pixel is chosen as median variance spectrum of the region surrounding that
pixel, thus avoiding to base the weight on the pixel variance. Slit-loss
correction has been applied on the combined spectrum based on the average seeing
of the observations. We show the extracted spectrum in Fig.~\ref{fig:SED}.

@@

We determine a redshift of $z = 2.211 \pm 0.001$ from the simultaneous detection
of emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959,
\oiii$\lambda$5007, and \ha. \hb{} and \oiii$\lambda$4959 are both detected at
low significance ($\sim$ 3-$\sigma$). The uncertainty on the redshift is the
standard deviation of independent measurements of the redshift based on the
individual lines. We show \oiii$\lambda$5007 in the insert in
Fig.~\ref{fig:line}.  \ha{} is only visible in the first epoch due to the
$K$-band blocking filter used for the remaining observations. The nebular lines
exhibit a spatial extent of $\sim$ 1\farc5 and show significant velocity
structure along the slit. A drop in the continuum bluewards of the \lya{} line
further supports the inferred redshift. No spectral evolution is observed across
the epochs indicating that there is negligible GRB afterglow contribution to the
spectra.

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of the
host \citep{Kennicutt1998}. At the redshift of the GRB host, \ha{} is observed
at 21\,000~\AA{} where the night sky is very bright. In addition, several bright
sky-lines are superposed on the line, making an accurate estimate of the
\ha-flux difficult. Due to their velocity structure, the lines exhibit clear
deviations from a Gaussian and given the low S/N of the spectra we do not
attempt any parametric fits. We instead obtain a limit on the SFR by numerically
integrating the part of \ha{} free of contamination and obtain $F_{H \alpha} >
4.1 \times 10^{-17} \mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2}$. After
converting the \citet{Kennicutt1998} relation to a \citet{Chabrier2003} initial
mass function using the conversion factor from \citet{Madau2014}, we derive a limit of $SFR > 7
M_\odot$~yr$^{-1}$. We additionally obtain the \lya{} line flux by numerically
integrating the entire \lya{} line complex and obtain $F_{\mathrm{Ly}\alpha} =
2.0 \pm 0.5 \times 10^{-17} \mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2}$,
where the error on the flux is found by integrating the associated error
spectrum over the same spectral region.

@@

From the SED-fit (Sect.~\ref{SED}), and the detection of \lya{}, the host is
constrained to contain very little or no dust, although the presence of \lya{}
does not exclude dust. Therefore we do not apply a dust-correction to the
measured \ha{} flux here. \oii{} is close to a region of strong telluric
absorption, which is why no SFR is inferred from this line.

The total extent of the lines in velocity space is $\sim$ 450 km/s. The line
profiles shows an asymmetric "double-horned" profile, indicating that we are
seeing a galaxy with a large degree of coherent rotational motion relative to
the line-of-sight. If we assume that we are viewing a spiral galaxy edge-on,
this is a measure of the rotational velocity of the gas. If we assume that the
spectral resolution and the turbulent width of the lines are negligible compared
to the rotational velocity, we can, based on the projected size of the source
and the width of the lines, put a constraint on the dynamical mass of the galaxy
\citep{DeBlok2014}. Based on the physical size along the slit and the velocity
width of \oiii$\lambda5007$, we infer M$_\text{dyn} \gtrsim 10^{10.8}$
M$_\odot$. Because we are viewing the host inclined at an angle relative to
edge-on and because the slit is not aligned along the long axis of the host,
this value is a lower limit.

@@

In addition to the spectroscopy presented above, we imaged the field of
GRB~111117A in multiple broad-band filters using the VLT equipped with FORS2
($gRIz$ filters) and HAWK-I ($JHK_{\mathrm{s}}$ filters), long after the burst
faded. These new data are complemented by a re-analysis of some of the imaging
used in \citet{Margutti2012} and \citet{Sakamoto2013} that are available to us
(GTC $gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in
Table~\ref{tab:phot_overview}. Most data were taken long after the GRB had faded
when no afterglow contribution was present. Given the faintness of the afterglow
(see Sect. \ref{xray}, and \citealt{Cucchiara2011, Cenko2011}), we also expect
negligible contribution to the earliest epochs, which is confirmed by the
consistency between the two $g$-band measurements.

All data were reduced, analyzed and fitted in a similar manner as described in
detail in \citet{Kruhler2011a} and, more recently, in \citet{Schulze2016}.
Briefly, we use our own \texttt{Python} and IRAF routines to perform a standard
reduction which includes bias/flat-field correction, de-fringing (where
necessary), sky-subtraction, and stacking of individual images. On the final
reduced image products, \texttt{DAOPHOT} \citep{Stetson1987} was used to derive
the reported photometry, where the size of the aperture was chosen to be
2\farc0. Because image quality ranges from 0\farc6 - 1\farc7 and the galaxy
major axis is $\sim$ 1\farc, this ensures that, in all cases, the large majority
of the flux is collected and low-surface brightness light missed in the aperture
will not influence the measurement. This method sacrifices some S/N in the best
seeing cases for more reliable photometry across differing observing conditions.

Photometric calibration was fixed relative to field stars from the SDSS and
2MASS catalogs in the case of $griz$ and $JHK_{\mathrm{s}}$ filters,
respectively. For the $R$- and $I$-band photometry, we used the color
transformations of Lupton%
\footnote{\url{https://www.sdss3.org/dr8/algorithms/sdssUBVRITransform.php}}. We
convert all magnitudes into the AB system, and correct for a Galactic foreground
of $E_{B-V}=0.027~\mathrm{mag}$ \citep{Schlegel1998, Schlafly2011}.

Noteworthy is the discrepancy of both our new VLT/FORS2 photometry and the
re-analysis of the Gemini data compared to the $z$-band measurements of
\citet{Margutti2012} and \citet{Sakamoto2013}. Both of these authors report
$z$-band magnitudes that are brighter by $\sim 1.0$ mag compared to our values,
where data taken in other filters are consistent within the errors. Visual
inspection of the Gemini image shows only a marginal detection of the host,
which we report here as a $\sim$ 2-$\sigma$ measurement. More conservatively,
the 3-$\sigma$ upper limit for the Gemini image is $z > 24.06$. Objects of
magnitude $z \sim 23$ are clearly seen and are significantly brighter than the
GRB host galaxy. The consistency between the deeper FORS2 $z$-band image and the
upper limit derived for the Gemini image lends credence to the inferred
magnitudes presented here, see Fig. \ref{fig:photplot}.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/photometry_mosaic.pdf}
	\caption{Mosaic showing all obtained imaging. The host of GRB~111117A is marked by a red circle with a 2\farc0 radius. This is the same size as the aperture used for deriving the photometry. Worth noting is the relative depth of the GMOS-N $z$-band and the FORS2 $z$-band. For reference, the object located to the south-east of the host and marked with a blue, dashed circle has an extinction corrected magnitude of $23.11 \pm 0.09$ ($23.10 \pm 0.18$) in the FORS2 (GMOS-N) $z$-band.}
	\label{fig:photplot}
\end{figure}

@@

We have here provided a revised, spectroscopic redshift measurement for the
short GRB~111117A based on host galaxy emission lines setting it at $z = 2.211
\pm 0.001$. This value supersedes the previous photometric redshift of
\citet{Margutti2012} and \citet{Sakamoto2013}. The erroneous redshift estimate
of previous authors is attributed to a discrepancy in the measured $z$-band
magnitude, and highlights the importance of deep spectroscopic studies of sGRB
hosts at medium resolution.

>>>>

We use the spectroscopic redshift, as well as new imaging data to constrain the
nature of the host galaxy and the physical parameters of the GRB. The rest-frame
X-ray derived hydrogen column density, for example, is the highest compared to a
complete sample of sGRBs and seems to follow the evolution with redshift as
traced by the hosts of long GRBs.

The host lies in the brighter end of the expected sGRB host brightness
distribution at $z = 2.211$, and is actively forming stars. Using the observed
sGRB host luminosity distribution, we find that between 43 and 71 per cent of
all \textit{Swift}-detected sGRBs have too faint hosts at $z\sim2$ to allow for
a secure redshift determination. This implies that the measured sGRB redshift
distribution could be incomplete at high redshift. The large $z$ of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts (sGRBs) come
from the merger of a neutron star (NS), either with another NS, or a black hole,
due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Berger2013b, Yang2015, Jin2016, Rosswog2016}. This association was
recently confirmed by the simultaneous and co-spatial detection of gravitational
waves (GWs) from a binary neutron star merger and a sGRB
\citep{LIGOScientificCollaboration2017a, Goldstein2017, Savchenko2017}. The
absence of associated supernovae in deep searches
\citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b, Kann2011} supports this idea and
distinguishes the physical origin of sGRBs from their long-duration
counterparts, \citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}.

The classification of GRBs in two groups initially comes from the bimodal
distribution of burst duration and spectral hardness in the BATSE sample \citep{Kouveliotou1993},
where the duration $T_{90} < 2$~s has been regarded as the dividing line between long
and short GRBs. Additionally, it has been found for long GRBs (lGRBs) that there is a
spectral lag in the arrival-time of photons, with the most energetic ones
arriving first. This lag is consistent with zero for sGRBs
\citep{Norris2006}. Because both populations have continuous, overlapping
distributions in their observables and because telescopes observe in differing
bands, it is difficult to impose a single demarcation criterion between the two
classes. For this reason, the distinction between long and short GRBs is
preferably based on a combination of the high-energy properties \citep{Zhang2009,
	Kann2011, Bromberg2012a, Bromberg2013}.

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}. Unlike lGRBs,
sGRB afterglows are faint making absorption spectroscopy often ineffective.
Therefore, most of these measurements come from the associated hosts and is
potentially biased towards lower redshifts. Additionally, because the
\textit{Swift} sensitivity to sGRB decreases sharply with redshift
\citep{Behroozi2014}, the intrinsic redshift distribution of sGRBs is largely
unknown at higher redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above the detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long lived in comparison to lGRBs, with the host stellar mass affecting
the sGRB rates more than the star-formation rate (SFR) \citep{Berger2014}.

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers \citep{Ghirlanda2016}. This epochal breakthrough occurred
recently when the first NS-NS GW event ever was detected by LIGO/Virgo (GW
170817) and associated to the weak sGRB 170817A detected by the \textit{Fermi}
and INTEGRAL satellites \citep{LIGOScientificCollaboration2017a, Goldstein2017,
	Savchenko2017}. The simultaneous detection of a sGRB and a GW provides new
promising ways to constrain the binary inclination angle \citep{Arun2014,
	LIGOScientificCollaboration2017a} and measure cosmological distances
\citep{Nissanke2010, LIGOScientificCollaboration2017c}.

The total lifetime of NS binaries depends on their orbit, mass, spin, initial
separations and subsequent inspiral times. The delay time from formation to
explosion impacts the timing and distribution of the enrichment of the ISM with
heavy $r$-process elements \citep{VandeVoort2015, Wallner2015,  Ji2016}. Some
limits can be calculated using host galaxy star-formation history models and
spatial distribution of sGRBs within their hosts \citep[][]{Berger2014}. The
most distant cosmological bursts, however, offer direct, hard limits on the
coalescence time scales.

We here present a spectrum of the host galaxy of the short GRB~111117A
($T_{90}=0.46$~s) and measure its redshift to be $z=2.211$. This value is
significantly higher than the previously estimated redshift based on photometric
studies \citep{Margutti2012,Sakamoto2013}. We present the GRB's rest frame
properties based on this new distance compared to previous analyses and revisit
the host properties derived from the new solution to the spectral energy
distribution (SED) fit. While no optical afterglow was detected, the excellent
localization from a detection of the X-ray afterglow by the \emph{Chandra} X-ray
Observatory allows us to discuss the positioning and environmental properties of
this remarkably distant sGRB. We use the $\Lambda$CDM cosmology parameters
provided by \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$. All magnitudes
are given in the AB system.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{
	FORS2 $R$-band imaging of the field of GRB~111117A with the X-shooter slit
overlaid. The slit position represents 4 epochs of spectroscopic observations
taken at similar position angles. The corresponding photometry is shown in
Fig.~\ref{fig:SED}. The blue asterisk indicates the GRB position as derived
from the \emph{Chandra} observations in \citet{Sakamoto2013}.
	}
	\label{fig:spec_setup}
\end{figure}

@@

\begin{figure*}
	\centering
	\includegraphics[width=18cm]{figures/grid.pdf}
	\caption{2D-images of the emission lines corresponding to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959, \oiii$\lambda$5007, and \ha. The location of bright skylines are marked by blue boxes. The locations of the emission lines are indicated with red ellipses. Because the host is observed in nodding-mode, negative images of the emission lines appear on both sides in the spatial direction. For the upper, left panel containing \lya, the systemic redshift position of \lya{} is marked by a yellow, vertical, dashed line. The red part of the \oii$\lambda$3727-doublet is affected by atmospheric absorption.}
	\label{fig:line}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=16cm]{figures/SEDspecphot.pdf}
	\caption{Best-fit SED to the derived photometry. The detection of \lya{}  is predicted from the SED fit and confirmed by the spectroscopic observations. Overplotted in grey is the observed spectrum, binned by 6 \AA{} for presentation purposes. Slit losses has been corrected for based on the average seeing of the observations, as confirmed by the comparison with the photometry. The blue, dashed line is the corresponding error spectrum, smoothed for presentation purposes. The reason for the spectral gaps at 5500 \AA{} and 10000 \AA{} is from the merging of the arms.}
	\label{fig:SED}
\end{figure*}

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). Observations use a simple ABBA nodding pattern, with 5
\arcsec~nod throws. X-shooter covers the wavelength range from 3000~\AA{} to
24\,800~\AA{} (21\,000~\AA{} when the $K$-band blocking filter is used) across
three spectroscopic arms. We carried out the bias-correction, flat-fielding,
order tracing, wavelength calibration, rectification, and flux calibration using
the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} run
in physical mode. Because the echelle orders are curved across each detector, a
rectification algorithm is employed which introduces correlations between
neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the
UVB/VIS/NIR arm to minimize the degree of correlation while conserving the
maximal resolution. The observations are combined and extracted using scripts
described in Selsing et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}},
where the full spectral point spread function is modeled across each arm and
used for the optimal extraction algorithm \citep{Horne1986}. An overview of the
spectroscopic observations is given in Table~\ref{tab:spec_overview}, and the
slit position is shown in Fig.~\ref{fig:spec_setup}. Each of the epochs are
extracted individually and combined in a weighted fashion where the weight at
each pixel is chosen as median variance spectrum of the region surrounding that
pixel, thus avoiding to base the weight on the pixel variance. Slit-loss
correction has been applied on the combined spectrum based on the average seeing
of the observations. We show the extracted spectrum in Fig.~\ref{fig:SED}.

@@

We determine a redshift of $z = 2.211 \pm 0.001$ from the simultaneous detection
of emission lines belonging to \lya, \oii$\lambda$3727, \hb, \oiii$\lambda$4959,
\oiii$\lambda$5007, and \ha. \oii$\lambda$3727, \hb{}, and \oiii$\lambda$4959
are detected at low significance ($\sim$ 3-$\sigma$). The uncertainty on the
redshift is the standard deviation of independent measurements of the redshift
based on the individual line centroids (excluding \lya). We show cutouts of the
2D spectrum at the position of all the detected lines in Fig.~\ref{fig:line}. 
\ha{} is only visible in the first epoch due to the $K$-band blocking filter
used for the remaining observations. The nebular lines exhibit a spatial extent
of $\sim$ 1\farc5 and show significant velocity structure along the slit. A drop
in the continuum bluewards of the \lya{} line further supports the redshift. No
spectral evolution is observed across the epochs indicating that there is
negligible GRB afterglow contribution to the first epoch spectrum.

Using the luminosity of \ha, we can infer the SFR of the host
\citep{Kennicutt1998}. At the redshift of the GRB host, \ha{} is observed at
around 21\,000~\AA{} where the night sky is very bright. In addition, several
bright sky-lines are superposed on the line, making an accurate estimate of the
\ha~flux difficult. Due to their velocity structure, the lines exhibit clear
deviations from a Gaussian and given the low S/N of the spectra we do not
attempt any parametric fits. We instead obtain a limit on the SFR by numerically
integrating the part of \ha{} free of contamination and obtain $F_{\mathrm{H,X} \alpha} >
4.1 \times 10^{-17}~\mathrm{erg}~\mathrm{s}^{-1}~\mathrm{cm}^{-2}$. After
converting the \citet{Kennicutt1998} relation to a \citet{Chabrier2003} initial
mass function using the conversion factor from \citet{Madau2014}, we derive a
limit of $\mathrm{SFR} > 7 M_\odot$~yr$^{-1}$. We additionally obtain the \lya{} line
flux by numerically integrating the entire \lya{} line complex and obtain
$F_{\mathrm{Ly}\alpha} = (2.0 \pm 0.5) \times 10^{-17}~\mathrm{erg}~
\mathrm{s}^{-1}~\mathrm{cm}^{-2}$, where the error on the flux is found by
integrating the associated error spectrum over the same spectral region.

@@

From the SED-fit (Sect.~\ref{SED}) the host is constrained to contain very
little or no dust. Consistently, \lya{} is detected although its presence does
not exclude dust. Therefore we do not apply a dust-correction to the measured
\ha{} flux here. \oii{} is close to a region of strong telluric absorption,
which is why no SFR is inferred from this line.

The total extent of the lines in velocity space is $\sim$ 450 km s$^{-1}$. The line
profiles shows an asymmetric "double-horned" profile, indicating that we are
seeing a galaxy with a large degree of coherent rotational motion relative to
the line-of-sight. If we assume that we are viewing a spiral galaxy edge-on,
this is a measure of the rotational velocity of the gas. If we assume that the
spectral resolution and the turbulent width of the lines are negligible compared
to the rotational velocity, we can, based on the projected size of the source
and the width of the lines, put a constraint on the dynamical mass of the galaxy
\citep{DeBlok2014}. Based on the physical size along the slit and the velocity
width of \oiii$\lambda5007$, we infer $M_\text{dyn} \gtrsim 10^{10.8}$
$M_\odot$. Because we are viewing the host inclined at an angle relative to
edge-on and because the slit is not aligned along the long axis of the host,
this value is a lower limit.

@@

In addition to the spectroscopy presented above, we imaged the field of
GRB~111117A in multiple broad-band filters using the VLT equipped with FORS2
($gRIz$ filters) and HAWK-I ($JHK_{\mathrm{s}}$ filters). These new data are
complemented by a re-analysis of some of the imaging used in
\citet{Margutti2012} and \citet{Sakamoto2013} that are available to us (GTC
$gri$-band, TNG $R$-band, and Gemini $z$-band). A log of the photometric
observations and measured brightnesses is given in
Table~\ref{tab:phot_overview}. Most data were taken long after the GRB had faded
when no afterglow contribution was present. Given the faintness of the afterglow
(see Sect. \ref{xray}, and \citealt{Cucchiara2011, Cenko2011}), we also expect
negligible contribution to the earliest epochs, which is confirmed by the
consistency between the two $g$-band measurements.

All data were reduced, analyzed and fitted in a similar manner as described in
detail in \citet{Kruhler2011a} and, more recently, in \citet{Schulze2016}. We
use our own \texttt{Python} and IRAF routines to perform a standard reduction
which includes bias/flat-field correction, de-fringing (where necessary),
sky-subtraction, and stacking of individual images. On the final reduced image
products, \texttt{DAOPHOT} \citep{Stetson1987} was used to derive the reported
photometry, where the size of the aperture was chosen to be 2\farc0. Because
image quality ranges from 0\farc6 - 1\farc7 and the galaxy major axis is $\sim$
1\farc, this ensures that, in all cases, the large majority of the flux is
collected and low-surface brightness light missed in the aperture will not
influence the measurement. This method sacrifices some S/N in the best seeing
cases for more reliable photometry across differing observing conditions.

Photometric calibration was fixed relative to field stars from the SDSS and
2MASS catalogs in the case of $griz$ and $JHK_{\mathrm{s}}$ filters,
respectively. For the $R$- and $I$-band photometry, we used the color
transformations of Lupton%
\footnote{\url{http://www.sdss3.org/dr8/algorithms/sdssUBVRITransform.php}}. We
convert all magnitudes into the AB system, and correct for a Galactic foreground
of $E_{B-V}=0.027~\mathrm{mag}$ \citep{Schlegel1998, Schlafly2011}.

Noteworthy is the discrepancy of both our new VLT/FORS2 photometry and the
re-analysis of the Gemini data compared to the $z$-band measurements of
\citet{Margutti2012} and \citet{Sakamoto2013}. Both of these authors report $z
\sim 23$, which is brighter than our measurement by $\sim 1.0$ mag, while data
taken in other filters are consistent within the errors. Visual inspection of
the Gemini image shows only a marginal detection of the host, which we report
here as a $\sim$ 2-$\sigma$ measurement. More conservatively, the 3-$\sigma$
upper limit for the Gemini image is $z > 24.06$. Objects of magnitude $z \sim
23$ are clearly seen and are significantly brighter than the GRB host galaxy.
The consistency between the deeper FORS2 $z$-band image and the upper limit
derived for the Gemini image lends credence to the inferred magnitudes presented
here, see Fig. \ref{fig:photplot}.

\begin{figure}
	\centering \includegraphics[width=9cm]{figures/photometry_mosaic.pdf}
\caption{Mosaic showing all used imaging. The host of GRB~111117A is marked by
	a red circle with a 2\farc0 radius. This is the same size as the aperture used
	for deriving the photometry. Each panel is 20\arcsec~in size, north is up, east
	is left. Worth noting is the relative depth of the GMOS-N and FORS2 $z$-band
	images. For reference, the object located to the south-east of the host and
	marked with a blue, dashed circle has an extinction corrected magnitude of
	$23.11 \pm 0.09$ ($23.10 \pm 0.18$) in the FORS2 (GMOS-N) $z$-band.}
\label{fig:photplot}
\end{figure}

@@

We have here provided a revised, spectroscopic redshift measurement for the
short GRB~111117A based on host galaxy emission lines setting it at $z = 2.211
\pm 0.001$. This value supersedes the previous photometric redshift estimates of
\citet{Margutti2012} and \citet{Sakamoto2013}. The erroneous best-fit SED
redshift of previous authors is attributed to a discrepancy in the measured
$z$-band magnitude, and highlights the importance of deep spectroscopic studies
of sGRB hosts at medium resolution.
